\pkg{webchem}: An \proglang{R} Package to retrieve Chemical Information from the Web

### 2016-04-23 revision1/article.Rnw
revised ms

\Plainauthor{Eduard Sz\"ocs, Ralf B. Sch\"afer}
\Plaintitle{webchem: an R Package to retrieve Chemical Information from the Web}
\Shorttitle{\pkg{webchem}: Chemical Information from the Web}

@@

options(prompt = "R> ", continue = "+  ", width = 80, useFancyQuotes = FALSE)
@

@@

When dealing with chemical datasets (e.g. environmental monitoring data, toxicological data), a first step is often to validate the names of chemicals or to link them to unique codes that simplify subsequent querying and appending of compound-related physico-chemical or toxicological information.
Several web sources provide chemical names or link them to unique codes (see also section \emph{Data sources} below).
However, manual searching for each compound, often through a graphical web interface, is tedious, error-prone and not reproducible \citep{Peng_2009}.

To simplify, robustify and automate this task, i.e. to search and retrieve chemical information from the web, we created the \pkg{webchem} package for the free and open source \proglang{R} language \citep{r_2015, Wehrens_2011}.
\proglang{R} is one of the most widely used software for data cleaning, analysing and visualising data, and supports full reproducibility of each step \citep{Marwick_2016}.

@@

Only few design decisions have been made:
Each function name has a prefix and suffix separated by an underscore \citep{Chamberlain_Szocs_2013}.
They follow the format of \code{source\_function}, e.g. \code{cs\_compinfo} uses ChemSpider as source (see next section) to retrieve compound information.
Some functions require querying first a unique identifier from the data source and then use this identifier to query further information.
The prefix \code{get} is used to denote these functions, e.g. \code{get\_csid} to retrieve the identifier used in ChemSpider.

@@

\section[Use Cases]{Use Cases}
\subsection[Install webchem]{Installation}
\pkg{webchem} can be easily installed and loaded from CRAN:

<<install, eval=FALSE>>=
install.packages("webchem")
library(webchem)
@

@@

(i) \code{jagst}: This dataset comprises environmental monitoring data of organic substances in the river Jagst of the year 2013, Germany.
The data is publicly available and can be retrieved from \citetalias{lubw_2016}.
It comprises concentrations  (in $\mu g~/~L$) of  34 substances  on 13 sampling occasions.

<<loadjagst>>=
# load jagst dataset
data(jagst)
# print first 6 lines of data
head(jagst)
@

@@

<<load_lc50>>=
data(lc50)
head(lc50)
@

@@

\subsection[Query Identifiers]{Query Identifiers}
The \code{jagst} data set covers 34 substances that are identified by (German) names.
Merging and linking these to other tables is hampered by differences and ambiguity in compound names.

One possibility to resolve this, is to use different chemical identifiers allowing easy identification.
There are several identifiers available, e.g.  registry numbers like CAS or EC, database identifiers like PubChemCID \citep{Kim_2016} or ChemSpiderID \citep{pence_chemspider:_2010}, line notations like SMILES \citep{Weininger_1990}, InChI and InChiKey \citep{Heller_McNaught_Pletnev_Stein_Tchekhovskoi_2015}. 
In this first example we query several identifiers to create a table that can be used as (i) supplemental information to a research article or (ii) to facilitate subsequent matching with other data.

@@

First we search for ETOX internal ID numbers:

<<get_etoxid, message=FALSE>>=
# unique substance names
subs <- unique(jagst$substance)
# search ETOX IDs
ids <- get_etoxid(subs, match = 'best')
head(ids)
@

@@

In the same manner we can now query other identifiers from other source using this CAS numbers (Fig. \ref{fig:fig1}):
<<token, echo=FALSE>>=
# this is a webchem specific token. 
# Please use it only for reproduction / testing.
token <- 'b44136c5-49cf-4df4-ba18-e4676449d1f4'
@

<<others, message = FALSE, warning=FALSE>>=
# query SMILES from PubChem
cids <- get_cid(etox_cas)
pc_data <- pc_prop(cids, properties = c('CanonicalSMILES'))
pc_smiles <- smiles(pc_data)

# query InChiKey from ChemSpider
# token <- '<yourtokenhere>'
csids <- get_csid(etox_cas, token = token)
cs_data <- cs_compinfo(csids, token = token)
cs_inchikey <- inchikey(cs_data)
@

<<table>>=
# combine in one data.frame
res <- data.frame(name = subs, cas = etox_cas, smiles = pc_smiles, 
                  cid = pc_data$CID, inchikey = cs_inchikey, 
                  csid = cs_data$csid, 
                  stringsAsFactors = FALSE)
@

Note, that in order to use the ChemSpider functions, a personal authentication key (\code{token}) is needed, which can be retrieved from the ChemSpider webpage. 
We end up with a compound table containing many different identifiers (Table~\ref{tab:comptable}), allowing easy identification and merging with other datasets, e.g. the lc50 dataset based on CAS.

\begin{table}[ht]
\centering
<<print_table, echo=FALSE, results = 'asis'>>=
tmp <- res[1:6, ]
tmp$smiles <- ifelse(is.na(tmp$smiles), NA, paste0(substr(tmp$smiles, 1, 7), '...'))
tmp$inchikey <- ifelse(is.na(tmp$inchikey), NA, paste0(substr(tmp$inchikey, 1, 7), '...'))
tmp[is.na(tmp)] <- '-'
colnames(tmp) <- c("Name", "CAS", "SMILES\\textsuperscript{a}","CID", "InChIKey\\textsuperscript{a}", "CSID")
print(xtable(tmp),
  floating = FALSE,
  hline.after = NULL,
  include.rownames = FALSE,
  add.to.row = list(pos = list(-1,0, 6),
                    command = c('\\toprule\n','\\midrule\n','\\bottomrule\n')),
  sanitize.text.function = function(x){x}
  )
@
\caption{Identifiers for the jagst datasets as queried with \pkg{webchem}. Only the first 6 entries are shown. - = not found. \textsuperscript{a} Only the first 7 characters are shown.}
\label{tab:comptable}
\end{table}

\subsection[Toxicity of different pesticide groups]{Toxicity of different pesticide groups}
Another question we might ask is \emph{How does toxicity vary between insecticide groups?}
Answering this question would require tedious lookup of insecticide groups for each of the 124 CAS numbers in the lc50 dataset.
The Compendium of Pesticide Common Names \citep{wood} contains such information and can be easily queried with \pkg{webchem}.

<<aw_query, message = FALSE>>=
# query the compendium using CAS-numbers
aw_data <- aw_query(lc50$cas, type = 'cas')
@

@@

<<subactivty, message = FALSE>>=
# shows internal structure of the data-object
# str(aw_data[[1]])
# extract chemical group
igroup <- sapply(aw_data, function(y) y$subactivity[1])
igroup[1:3]
@

@@

\begin{figure}[ht]
<<plot_lc50, echo=FALSE, out.width='0.7\\textwidth', fig.height=6>>=
# cleanup
lc50$type <- ifelse(grepl('carbamat', igroup), 'Carbamates',
              ifelse(grepl('neonicotinoid', igroup), 'Neo-\nnicotinoids', 
                ifelse(grepl('pyrethroid', igroup), 'Pyrethroids', 
                  ifelse(grepl('organophosphate', igroup), 'Organo-\nphosphates', 
                         'other'))))
lc50$type <- factor(lc50$type, levels = c("Pyrethroids", "Carbamates", "Organo-\nphosphates", "Neo-\nnicotinoids", "other"))

# plot
p <- ggplot(lc50, aes(x = type, y = value)) +
  geom_boxplot(fill = 'grey75') + 
  geom_jitter(width = 0.5) +
  scale_y_log10() +
  labs(y = expression(LC[50~'D.magna, 48h']), x = '') +
  theme_bw() +
  theme(text = element_text(size = 16))
p
@
\caption{Toxicity of different pesticide groups. LC50 values have been retrieved from EPA ECOTOX database, chemical groups from the Compendium of Pesticide Common Names \citep{wood}}
\label{fig:fig2}
\end{figure}

@@

<<qsar, message=FALSE, cache=TRUE, warning=FALSE>>=
# query PHYSPROP DATABASE
pp_data <- pp_query(lc50$cas)
@

@@

<<qsar2, warnings = FALSE>>=
# look  at internal structure
# str(pp_data[[1]])
lc50$logp <- sapply(pp_data, function(y){
  if (length(y) == 1 && is.na(y))
    return(NA)
  y$prop$value[y$prop$variable == 'Log P (octanol-water)']
})
@

@@

\subsection[Utility functions]{Utility functions}
\pkg{webchem} provides also basic functions to check identifiers that can be used for data quality assessment.
The functions either use simple formatting rules or web resources.

<<check>>=
# simple formatting check
is.inchikey('BQJCRHHNABKAKU-KBQPJGBKS-AN')
# or using ChemSpider
is.inchikey('BQJCRHHNABKAKU-KBQPJGBKSA-5', 
      type = 'chemspider')
# formatting check
is.cas('64-17-6')
@

\section[Discussion]{Discussion}
\subsection[Related software]{Related software}
Within the \proglang{R} ecosystem , there are only few similar projects:
\pkg{rpubchem} \citep{rpubchem_2014} and \pkg{ChemmineR} \citep{chemminer_2008} both provide an interface to PubChem that is more extensive than the current one in \pkg{webchem}.
\pkg{WikidataR} \citep{wikidatar_2016} provides an interface to wikidata that could be used to retrieve chemical data from Wikipedia.
However, it does not provide predefined methods for chemical data like \pkg{webchem}.
Within the \proglang{Python} ecosystem (a popular programming language) the libraries \pkg{PubChempy} \citep{pubchempy}, \pkg{ChemSpiPy} \citep{chemspipy} and \pkg{CIRpy} \citep{cirpy} are available for similar tasks as those outlined here.
\pkg{webchem} is not specialized and tries to integrate many data sources and for some of these provides a unique programmatic interface.
The Chemical Translation Service \citep{Wohlgemuth_Haldiya_Willighagen_Kind_Fiehn_2010}, which is also one of the sources that can be queried,  allows batch conversion of chemical identifiers.
However, it does not provide access to other data (experimental, modeled or legal data).

>>>>

\Plainauthor{Eduard Sz\"ocs, Ralf B. Sch\"afer}
\Plaintitle{webchem: an R Package to Retrieve Chemical Information from the Web}
\Shorttitle{\pkg{webchem}: Chemical Information from the Web}

@@

options(prompt = "R> ", continue = "+  ", width = 76, useFancyQuotes = FALSE)
@

@@

When dealing with chemical datasets (e.g.\ environmental monitoring data, toxicological data), a first step is often to validate the names of chemicals or to link them to unique codes that simplify subsequent querying and appending of compound-related physico-chemical or toxicological information.
Several web sources provide chemical names or link them to unique codes (see also section \emph{Data sources} below).
However, manual searching for each compound, often through a graphical web interface, is tedious, error-prone and not reproducible \citep{Peng_2009}.

To simplify, robustify and automate this task, i.e.\ to search and retrieve chemical information from the web, we created the \pkg{webchem} package for the free and open source \proglang{R} language \citep{r_2015, Wehrens_2011}.
\proglang{R} is one of the most widely used software for data cleaning, analysing and visualising data, and supports full reproducibility of each step \citep{Marwick_2016}.

@@

Only few design decisions have been made:
Each function name has a prefix and suffix separated by an underscore \citep{Chamberlain_Szocs_2013}.
They follow the format of \code{source\_function}, e.g.\ \code{cs\_compinfo} uses ChemSpider as source (see next section) to retrieve compound information.
Some functions require querying first a unique identifier from the data source and then use this identifier to query further information.
The prefix \code{get} is used to denote these functions, e.g.\ \code{get\_csid} to retrieve the identifier used in ChemSpider.

@@

\section[Use cases]{Use cases}
\subsection[Install webchem]{Installation}
\pkg{webchem} can be easily installed and loaded from CRAN:

<<install, eval=FALSE>>=
install.packages("webchem")
library("webchem")
@

@@

(i) \code{jagst}: This dataset comprises environmental monitoring data of organic substances in the river Jagst of the year 2013, Germany.
The data is publicly available and can be retrieved from \citetalias{lubw_2016}.
It comprises concentrations  (in $\mu g~/~L$) of  34 substances  on 13 sampling occasions.
First we load the dataset and take a look at the first six rows:

<<loadjagst>>=
data("jagst")
head(jagst)
@

@@

<<load_lc50>>=
data("lc50")
head(lc50)
@

@@

\subsection[Query identifiers]{Query identifiers}
The \code{jagst} data set covers 34 substances that are identified by (German) names.
Merging and linking these to other tables is hampered by differences and ambiguity in compound names.

One possibility to resolve this, is to use different chemical identifiers allowing easy identification.
There are several identifiers available, e.g.\  registry numbers like CAS or EC, database identifiers like PubChemCID \citep{Kim_2016} or ChemSpiderID \citep{pence_chemspider:_2010}, line notations like SMILES \citep{Weininger_1990}, InChI and InChiKey \citep{Heller_McNaught_Pletnev_Stein_Tchekhovskoi_2015}. 
In this first example we query several identifiers to create a table that can be used as (i) supplemental information to a research article or (ii) to facilitate subsequent matching with other data.

@@

First we search for ETOX internal ID numbers using the substance names:

<<get_etoxid, message=FALSE, cache=TRUE>>=
subs <- unique(jagst$substance)
ids <- get_etoxid(subs, match = 'best')
head(ids)
@

@@

In the same manner we can now query other identifiers from other source using this CAS numbers (Fig. \ref{fig:fig1}), like PubChem
<<token, echo=FALSE>>=
# this is a webchem specific token. 
# Please use it only for reproduction / testing.
token <- 'b44136c5-49cf-4df4-ba18-e4676449d1f4'
@

<<pubchem, message = FALSE, warning=FALSE>>=
cids <- get_cid(etox_cas)
pc_data <- pc_prop(cids, properties = c('CanonicalSMILES'))
pc_smiles <- smiles(pc_data)
@

or ChemSpider
<<chemspider, message = FALSE, warning=FALSE>>=
csids <- get_csid(etox_cas, token = token)
cs_data <- cs_compinfo(csids, token = token)
cs_inchikey <- inchikey(cs_data)
@

Finally, we combine the queried data into one data.frame
<<table>>=
res <- data.frame(name = subs, cas = etox_cas, smiles = pc_smiles, 
  cid = pc_data$CID, inchikey = cs_inchikey, csid = cs_data$csid, 
  stringsAsFactors = FALSE)
@

Note, that in order to use the ChemSpider functions, a personal authentication key (\code{token}) is needed, which can be retrieved from the ChemSpider webpage. 
We end up with a compound table containing many different identifiers (Table~\ref{tab:comptable}), allowing easy identification and merging with other datasets, e.g.\ the lc50 dataset based on CAS.

\begin{table}[ht]
\centering
<<print_table, echo=FALSE, results = 'asis'>>=
tmp <- res[1:6, ]
tmp$smiles <- ifelse(is.na(tmp$smiles), NA, paste0(substr(tmp$smiles, 1, 7), '...'))
tmp$inchikey <- ifelse(is.na(tmp$inchikey), NA, paste0(substr(tmp$inchikey, 1, 7), '...'))
tmp[is.na(tmp)] <- '-'
colnames(tmp) <- c("Name", "CAS", "SMILES","CID", "InChIKey", "CSID")
print(xtable(tmp),
  floating = FALSE,
  hline.after = NULL,
  include.rownames = FALSE,
  add.to.row = list(pos = list(-1,0, 6),
                    command = c('\\toprule\n','\\midrule\n','\\bottomrule\n')),
  sanitize.text.function = function(x){x}
  )
@
\caption{Identifiers for the jagst datasets as queried with \pkg{webchem}. Only the first 6 entries are shown. For SMILES and InChIKey only the first 7 characters are shown. - = not found.}
\label{tab:comptable}
\end{table}

\subsection[Toxicity of different pesticide groups]{Toxicity of different pesticide groups}
Another question we might ask is \emph{How does toxicity vary between insecticide groups?}
Answering this question would require tedious lookup of insecticide groups for each of the 124 CAS numbers in the lc50 dataset.
The Compendium of Pesticide Common Names \citep{wood} contains such information and can be easily queried using CAS numbers with \pkg{webchem}:

<<aw_query, message = FALSE>>=
aw_data <- aw_query(lc50$cas, type = 'cas')
@

@@

<<subactivty, message = FALSE>>=
igroup <- sapply(aw_data, function(y) y$subactivity[1])
igroup[1:3]
@

@@

\begin{figure}[ht]
<<plot_lc50, echo=FALSE, out.width='0.7\\textwidth', fig.height=6>>=
lc50$type <- ifelse(grepl('carbamat', igroup), 'Carbamates',
              ifelse(grepl('neonicotinoid', igroup), 'Neo-\nnicotinoids', 
                ifelse(grepl('pyrethroid', igroup), 'Pyrethroids', 
                  ifelse(grepl('organophosphate', igroup), 'Organo-\nphosphates', 
                         'other'))))
lc50$type <- factor(lc50$type, levels = c("Pyrethroids", "Carbamates", "Organo-\nphosphates", "Neo-\nnicotinoids", "other"))

p <- ggplot(lc50, aes(x = type, y = value)) +
  geom_boxplot(fill = 'grey75') + 
  geom_jitter(width = 0.5) +
  scale_y_log10() +
  labs(y = expression(LC[50~'D.magna, 48h']), x = '') +
  theme_bw() +
  theme(text = element_text(size = 16))
p
@
\caption{Toxicity of different pesticide groups. LC50 values have been retrieved from EPA ECOTOX database, chemical groups from the Compendium of Pesticide Common Names \citep{wood}}
\label{fig:fig2}
\end{figure}

@@

<<qsar, message=FALSE, cache=TRUE, warning=FALSE>>=
pp_data <- pp_query(lc50$cas)
@

@@

<<qsar2, warnings = FALSE>>=
lc50$logp <- sapply(pp_data, function(y) {
  if (length(y) == 1 && is.na(y))
    return(NA)
  y$prop$value[y$prop$variable == 'Log P (octanol-water)']
})
@

@@

\subsection[Utility functions]{Utility functions}
\pkg{webchem} provides also basic functions to check identifiers that can be used for data quality assessment.
The functions either use simple formatting rules,

<<check_format>>=
is.inchikey('BQJCRHHNABKAKU-KBQPJGBKS-AN')
is.cas('64-17-6')
@

or web resources like ChemSpider
<<check_web>>=
is.inchikey('BQJCRHHNABKAKU-KBQPJGBKSA-5', 
  type = 'chemspider')
@

\section[Discussion]{Discussion}
\subsection[Related software]{Related software}
Within the \proglang{R} ecosystem , there are only few similar projects:
\pkg{rpubchem} \citep{rpubchem_2014} and \pkg{ChemmineR} \citep{chemminer_2008} both provide an interface to PubChem that is more extensive than the current one in \pkg{webchem}.
\pkg{WikidataR} \citep{wikidatar_2016} provides an interface to wikidata that could be used to retrieve chemical data from Wikipedia.
However, it does not provide predefined methods for chemical data like \pkg{webchem}.
Within the \proglang{Python} ecosystem the libraries \pkg{PubChempy} \citep{pubchempy}, \pkg{ChemSpiPy} \citep{chemspipy} and \pkg{CIRpy} \citep{cirpy} are available for similar tasks as those outlined here.
\pkg{webchem} is not specialized and tries to integrate many data sources and for some of these provides a unique programmatic interface.
The Chemical Translation Service \citep{Wohlgemuth_Haldiya_Willighagen_Kind_Fiehn_2010}, which is also one of the sources that can be queried,  allows batch conversion of chemical identifiers.
However, it does not provide access to other data (experimental, modeled or legal data).
